Recursive Solutions to First-Order Model Counting

### 2022-03-16 aiai_talk/talk.tex
First draft of a full talk

\documentclass{beamer}
\usepackage[utf8]{inputenc}
\usepackage[UKenglish]{babel}
\usepackage[UKenglish]{isodate}
\usepackage{amsmath}
\usepackage{mathtools}
\usepackage{tikz}

@@

\usetikzlibrary{shapes}
\usetikzlibrary{arrows}
\usetikzlibrary{arrows.meta}
\usetikzlibrary{positioning}

@@

@@

\begin{frame}{Some Elementary Counting}
  \begin{block}{A Counting Problem}
    Suppose we were meeting in person, the room had \structure{$n$} seats, and there were \structure{$m \le n$} attendees. How many ways would there be to seat everyone?
  \end{block}

@@

\begin{frame}{Overview of the Problem}
  \begin{itemize}
  \item \alert{First-order model counting} is the problem of counting the models of a sentence in first-order logic.
  \item The \alert{(symmetric) weighted} variation of the problem adds weights (e.g., probabilities) to predicates.
  \item Thus, SWFOMC can also be used for efficient \alert{probabilistic inference} in relational models.
  \item None of the (implemented) (SW)FOMC algorithms are able to count, e.g., \alert{injective} and \alert{bijective} functions.
  \end{itemize}
\end{frame}

% TODO: this can be fixed via recursion. Add a numerical example, show its complexity.
\begin{frame}{Recursion to the Rescue!}
  \[
  f(m, n) =
  \begin{cases}
    1 & \text{if } m = 0 \text{ and } n = 0 \\
    0 & \text{if } m > 0 \text{ and } n = 0 \\
    f(m, n-1) + mf(m-1, n-1) & \text{otherwise.}
  \end{cases}
  \]
  \begin{itemize}
  \item Optimal time complexity to compute \structure{$n^{\underline{m}}$} is \structure{$\Theta(\log m)$}
  \end{itemize}
\end{frame}

>>>>

\documentclass{beamer}
\usepackage[utf8]{inputenc}
\usepackage[UKenglish]{babel}
\usepackage[UKenglish]{isodate}
\usepackage{amsmath}
\usepackage{mathtools}
\usepackage{tikz}
\usepackage{complexity}

@@

\usetikzlibrary{shapes}
\usetikzlibrary{arrows}
\usetikzlibrary{arrows.meta}
\usetikzlibrary{positioning}
\usetikzlibrary{overlay-beamer-styles}

@@

\definecolor{color1}{HTML}{1B9E77}
\definecolor{color2}{HTML}{D95F02}

@@

\section{A Motivating Example}

\begin{frame}{Some Elementary Counting}
  \begin{exampleblock}{A Counting Problem}
    Suppose we were meeting in person, the room had \structure{$n$} seats, and there were \structure{$m \le n$} attendees. How many ways would there be to seat everyone?
  \end{exampleblock}

@@

\begin{frame}{Overview of the Problem}
  \begin{itemize}
  \item \alert{First-order model counting} is the problem of counting the models of a sentence in first-order logic.
  \item The \alert{(symmetric) weighted} variation of the problem adds weights (e.g., probabilities) to predicates.
  \item Thus, SWFOMC can also be used for efficient \alert{probabilistic inference} in relational models.
  \item None of the (implemented) (SW)FOMC algorithms are able to count, e.g., \alert{injective} and \alert{bijective} functions.
  \end{itemize}
  \begin{block}{Claim}
    This shortcoming can be addressed via support for (almost arbitrary) \alert{recursive functions}.
  \end{block}
\end{frame}

\begin{frame}{Back to Our Example}
  For instance, the following function counts injections
  \[
  f(m, n) =
  \begin{cases}
    1 & \text{if } m = 0 \text{ and } n = 0 \\
    0 & \text{if } m > 0 \text{ and } n = 0 \\
    f(m, n-1) + mf(m-1, n-1) & \text{otherwise.}
  \end{cases}
  \]
  \pause
  \begin{itemize}
  \item \structure{$f$} can be computed in \structure{$\Theta(mn)$} time (via dynamic programming).
  \item Optimal time complexity to compute \structure{$n^{\underline{m}}$} is \structure{$\Theta(\log m)$}.
  \item But \structure{$\Theta(mn)$} is still much better than solving an equivalent \alert{$\#\P$-complete} problem in propositional logic.
  \item The rest of this talk is about how such functions can be found automatically.
  \end{itemize}
\end{frame}

\section{Graphs and Their Role in Knowledge Compilation}

\begin{frame}{First-Order Knowledge Compilation}
  \begin{columns}
    \begin{column}{0.25\textwidth}
      \centering
      \begin{tikzpicture}[every node/.style={draw,ellipse},edge from parent/.style={draw,-latex},sibling distance=10mm,level distance=10mm]
        \node[fill=red!50,fill on=<2>] (dr) {}
        child {node {}
          child {node {}
            child {node {}}
            child {node {}
              child {node {}}
              child {node {}
                child {node {}}
                child {node[fill=red!50,fill on=<2>] {}
                  child {node {}
                    child {node {}}
                    child {node[fill=red!50,fill on=<2>] (ref) {}}
        }}}}}};
        \draw[-latex, bend right,alt=<2>{red}{black}] (ref) to (dr);
      \end{tikzpicture}
    \end{column}
    \begin{column}{0.75\textwidth}
      \begin{block}{Workflow Before}
        \begin{enumerate}
        \item Compile the formula to a \alert{circuit}
        \item Evaluate to get the answer
        \end{enumerate}
      \end{block}
      \pause
      \begin{block}{Workflow After}
        \begin{enumerate}
        \item Compile the formula to a \alert{graph}
        \item Extract the definitions of functions
        \item Simplify
        \item Supplement with \alert{base cases}
        \item Evaluate to get the answer
        \end{enumerate}
      \end{block}
    \end{column}
  \end{columns}
\end{frame}
% Only Step 1 is fully automated at this point. The other steps are executed by hand.
% Two parts: compilation and inference/evaluation.

\begin{frame}{More Formally...}
  \begin{definition}
    A \alert{first-order deterministic decomposable negation normal form computational graph} (FCG) is a
    \begin{itemize}
    \item directed graph
    \item (which is weakly connected)
    \item with a single source,
    \item labelled vertices,
    \item and ordered outgoing edges.
    \end{itemize}
  \end{definition}
\end{frame}

\begin{frame}{How to Interpret an FCG}
  \begin{columns}
    \begin{column}{0.25\textwidth}
      \centering
      \begin{tikzpicture}[every node/.style={draw,ellipse},edge from parent/.style={draw,-latex},sibling distance=10mm,level distance=10mm]
        \node[fill=red!50,fill on=<2>] (dr) {}
        child {node[fill=red!50,fill on=<3>] {}
          child {node[fill=red!50, fill on=<4>] {}
            child {node[fill=red!50, fill on=<5>] {}}
            child {node[fill=red!50, fill on=<6>] {}
              child {node[fill=red!50, fill on=<7>] {}}
              child {node[fill=red!50, fill on=<8>] {}
                child {node[fill=red!50, fill on=<9>] {}}
                child {node[fill=red!50, fill on=<10>] {}
                  child {node[fill=red!50, fill on=<11>] {}
                    child {node[fill=red!50, fill on=<12>] {}}
                    child {node[fill=red!50, fill on=<13>] (ref) {}}
        }}}}}};
        \draw[-latex, bend right] (ref) to (dr);
      \end{tikzpicture}
    \end{column}
    \begin{column}{0.75\textwidth}
      \begin{align*}
        f(m, n) &= \onslide<3->{\sum_{l=0}^m \binom{m}{l}} \onslide<5->{1} \onslide<4->{\times} \onslide<7->{1} \onslide<6->{\times} \onslide<9->{1} \onslide<8->{\times} \onslide<12->{[l<2]} \onslide<11->{\times} \onslide<13->{f(m-l, n-1)}\\
        \onslide<14>{&= f(m, n-1) + mf(m-1, n-1)}
      \end{align*}
      \onslide<12>{
        \[
          [\phi] =
          \begin{cases}
            1 & \text{if } \phi \\
            0 & \text{if } \neg\phi
          \end{cases}
          \]
      }
    \end{column}
  \end{columns}
\end{frame} % Things get more interesting if you do some computations before encountering a cycle or if you have several cycles.

%% \section{Compilation: How to Construct an FCG}
\section{Compilation: How to Construct an FCG}

\begin{frame}{Compilation Rules}
  \begin{definition}
    A \alert{(compilation) rule} is a function that takes a \alert{formula} and returns a set of \structure{$(G, L)$} pairs, where
    \begin{itemize}
    \item \structure{$G$} is a (potentially \texttt{null}) FCG,
    \item and \structure{$L$} is a list of formulas.
    \end{itemize}
  \end{definition}
\end{frame}

\begin{frame}{Example Rule: Independence}
  Input formula:
  \begin{gather}
    ({\color{color1} \forall x, y \in L. x=y}) \land \label{eq:1} \tag{{\color{color1} 1}} \\
    ({\color{color2} \forall x \in M. \forall y, z \in N. P(x, y) \land P(x, z) \Rightarrow y=z}) \land \label{eq:2} \tag{{\color{color2} 2}} \\
    ({\color{color2} \forall w, x \in M. \forall y \in N. P(w, y) \land P(x, y) \Rightarrow w=x}) \label{eq:3} \tag{{\color{color2} 3}}
  \end{gather}
  \pause
  Only one \structure{(G, L)} pair:
  \[
  G = 
  \begin{tikzpicture}[edge from parent/.style={draw,-latex},baseline=(current bounding box.center)]
    \node[draw,circle] {$\land$}
    child {node {\color{color1} $\star$}}
    child {node {\color{color2} $\star$}}
    ;
  \end{tikzpicture},
  \qquad
  L = \langle {\color{color1} \eqref{eq:1}}, {\color{color2} \eqref{eq:2} \land \eqref{eq:3}} \rangle
  \]
  Later, both \structure{$G$} and \structure{$L$} are incorporated into a larger \alert{search state}.
\end{frame}
% stars mark edges to nowhere and they are associated with the formulas in L via a certain order

\begin{frame}{New Rule 1: (Generalised) Domain Recursion}
  \begin{columns}
    \begin{column}{0.20\textwidth}
      \centering
      \begin{tikzpicture}[every node/.style={draw,ellipse},edge from parent/.style={draw,-latex},sibling distance=10mm,level distance=10mm]
        \node[fill=red!50] (dr) {}
        child {node {}
          child {node {}
            child {node {}}
            child {node {}
              child {node {}}
              child {node {}
                child {node {}}
                child {node {}
                  child {node {}
                    child {node {}}
                    child {node (ref) {}}
        }}}}}};
        \draw[-latex, bend right] (ref) to (dr);
      \end{tikzpicture}
    \end{column}
    \begin{column}{0.8\textwidth}
      \begin{example}
        Input formula:
        \[
        \forall x \in M. \forall y, z \in N. y \ne z \Rightarrow \neg P(x, y) \lor \neg P(x, z)
        \]
        Output formula (with a new constant \structure{$c \in M$}):
        \begin{gather*}
          \forall y, z \in N. y \ne z \Rightarrow \neg P(\alert{c}, y) \lor \neg P(\alert{c}, z) \\
          \forall x \in M. \forall y, z \in N. \alert{x \ne c} \land y \ne z \Rightarrow \neg P(x, y) \lor \neg P(x, z)
        \end{gather*}
        %% Output FCG:
        %% \begin{tikzpicture}[edge from parent/.style={draw,-latex}]
        %%   \node[draw,circle] {$\textsc{DR}$}
        %%   child {node {$\star$}}
        %%   ;
        %% \end{tikzpicture}
        Here, domain recursion is applied to domain \structure{$M$}. It could similarly be applied to \structure{$N$} as well.
      \end{example}
    \end{column}
  \end{columns}
\end{frame}
% We're still creating a vertex in a graph with one outgoing edge.

\begin{frame}{New Rule 2: Constraint Removal}
  \begin{columns}
    \begin{column}{0.20\textwidth}
      \centering
      \begin{tikzpicture}[every node/.style={draw,ellipse},edge from parent/.style={draw,-latex},sibling distance=10mm,level distance=10mm]
        \node (dr) {}
        child {node {}
          child {node {}
            child {node {}}
            child {node {}
              child {node {}}
              child {node {}
                child {node {}}
                child {node[fill=red!50] {}
                  child {node {}
                    child {node {}}
                    child {node (ref) {}}
        }}}}}};
        \draw[-latex, bend right] (ref) to (dr);
      \end{tikzpicture}
    \end{column}
    \begin{column}{0.8\textwidth}
      \begin{example}
        Input formula (with a constant \structure{$c \in M$}):
        \begin{gather*}
        \forall x \in \alert{M}. \forall y, z \in N. \alert{x \ne c} \land y \ne z \Rightarrow \neg P(x, y) \lor \neg P(x, z) \\
        \begin{multlined}
          \forall w,x \in \alert{M}. \forall y \in N. \alert{w \ne c} \land \alert{x \ne c} \land w \ne x \Rightarrow \\
          \neg P(w, y) \lor \neg P(x, y)
        \end{multlined}
        \end{gather*}
        Output formula (with a new domain \structure{$M' \coloneqq M \setminus \{\, c \,\}$}):
        \begin{gather*}
          \forall x \in \alert{M'}. \forall y, z \in N. y \ne z \Rightarrow \neg P(x, y) \lor \neg P(x, z) \\
          \forall w,x \in \alert{M'}. \forall y \in N. w \ne x \Rightarrow \neg P(w, y) \lor \neg P(x, y)
        \end{gather*}
      \end{example}
    \end{column}
  \end{columns}
\end{frame}
% Requirement: all variables from M are != c.
% Necessary because later we want to compare formulas and identify that they are isomorphic up to domains.

\begin{frame}{New Rule 3: Identifying Possibilities for Recursion}
  \begin{columns}
    \begin{column}{0.25\textwidth}
      \centering
      \begin{tikzpicture}[every node/.style={draw,ellipse},edge from parent/.style={draw,-latex},sibling distance=10mm,level distance=10mm]
        \node (dr) {}
        child {node {}
          child {node {}
            child {node {}}
            child {node {}
              child {node {}}
              child {node {}
                child {node {}}
                child {node {}
                  child {node {}
                    child {node {}}
                    child {node[fill=red!50] (ref) {}}
        }}}}}};
        \draw[-latex, bend right, red] (ref) to (dr);
      \end{tikzpicture}
    \end{column}
    \begin{column}{0.75\textwidth}
      \begin{block}{Goal}
      Check if the input formula is isomorphic (up to domains) to a previously encountered formula.
      \end{block}
      \begin{block}{Rough Outline}
        \begin{enumerate}
        \item Consider pairs of `similar' clauses.
        \item Consider bijections between their sets of variables.
        \item Extend each such bijection to a map between sets of domains.
        \item If the bijection makes the clauses equal, and the domain map is compatible with previous domain maps, move on to another pair of clauses.
        \end{enumerate}
      \end{block}
    \end{column}
  \end{columns}
\end{frame}

%% \begin{frame}{Compilation as Search}
%%   \begin{definition}
%%     A \alert{(search) state} is a tuple \structure{$(G, C, L)$}, where:
%%     \begin{itemize}
%%     \item \structure{$G$} is an FCG (or \texttt{null}),
%%     \item \structure{$C$} is a compilation cache that maps integers to sets of pairs $(\phi, v)$, where $\phi$ is a formula, and $v$ is a vertex of $G$ (which is used to identify opportunities for recursion),
%%     \item and \structure{$L$} is a list of formulas (that are yet to be compiled). (Note that the order is crucial!)
%%     \end{itemize}
%%   \end{definition}
%%   The search algorithm combines greedy and breadth-first search:
%% \end{frame}

\section{Summary}

\begin{frame}{Summary \& Future Work}
  \begin{block}{Summary}
    The circuits hitherto used for FOMC become more powerful with:
    \begin{itemize}
    \item cycles,
    \item generalised domain recursion,
    \item and some more new compilation rules that support domain recursion.
    \end{itemize}
  \end{block}
  \begin{block}{Future Work}
    \begin{itemize}
    \item Automate:
      \begin{itemize}
      \item extracting and simplifying the definitions of functions,
      \item finding all base cases.
      \end{itemize}
    \item Open questions:
      \begin{itemize}
      \item What kind of \alert{sequences} are computable in this way?
      \item Would using a \alert{different logic} extend the capabilities of FOMC further?
      \end{itemize}
    \end{itemize}
  \end{block}
\end{frame}
